---
title: "Spatially-explicit model of cosmogenic nuclide production in sediments"
author: "Nicolas Gauthier and Nari Miller"
date: "October 20, 2016"
output:
  html_document: 
    keep_md: yes
  pdf_document:
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=F, warning=F}
library(magrittr)
library(raster)
library(ggplot2)
library(reshape2)
```

## Model specification
We use the following model for $P_z$, the production rate of $^{10}$Be at depth $z$:  
$$P_z = P_0 e^{-z\frac{L}{p}} - N\lambda$$  
with parameters:  

$P_0$ is the production rate of Beryllium-10 in quartz at sea level
```{r}
P_0 <- 4.49		# [at/g/yr] after Stone, 1999.
```

$\lambda$ is the decay constant for $^{10}$Be
```{r}
ltlambda <- log(2) / 1.5e6
```

$L$ is the absorption mean-free path (attenuation length)
```{r}
L <- 160 	# [g/cm2]
```  

$p$ is the density of overburden
```{r}
p <- 2.6 	# [ g/cm3]
```  

and variables:

$N$ is the concentration of nuclides in the sample  
$z$ is the depth to that packet of sediment at time $t$   
$t$ is some length of time.  

### Assumptions  
For the sake of simplicity, we assume there is no topographic shielding (topographic shielding factor = 1) and a constant location in the Mediterranean at 40N 0E.


## Sample data
First we need base set of raster maps to initialize the variables.  
We create a multi-layer raster brick where each cell represents a 1 cm$^{3}$ packet of sediment, the **values()** of the cells correspond to $N$ (the concentration of nuclides in that packet), and the index of each layer corresponds to $1 + z$, the depth of that packet of sediment.  

Lets create a sample raster brick with 5 layers of 1x1 cells, with an initial value of $N = 1000$ for each cell.  
```{r}
rast <- matrix(1000) %>% raster 
N0 <- brick(c(rast, rast, rast, rast, rast, rast, rast, rast, rast, rast))
```

## Simulation
First translate the above formula for $P_z$ into an R function that calculates the $^{10}$Be production rate given values of $N$ and $z$.
```{r}
P_z <- function(N, z){
  P_0 * exp(-z*L/p) - N*ltlambda
}
```

Numerically integrate the differential equation with Euler's method. Using this function and the sample raster brick, iterate over a period of 100 years.
```{r}
nsim <- 100 # simulation length

N <- N0 # initial conditions

record <- values(N) # vector to store the outputs of the simulation

for(i in 1:nsim){
  delta <- P_z(N, 1:nlayers(N) - 1)
  N <- N + delta
  record <- rbind(record, values(N))
}

```

Plot the resulting solution.
```{r}
ggplot(melt(record), aes(x=Var1,y=value, color = Var2)) + geom_line()
```

Note that the sediment packets at depth are decaying with time, just at a rate small enough not to be visible when compared with the rate of change of the surface level.

## Numerical integration
Maybe euler's method is introducing some errors, try a more advanced ODE solver? Still in progress ...

Redefine the function to be deSolve compataible.
```{r}
library(deSolve)
library(phaseR)

be10 <- function(t, y, parameters){
    z <- parameters
    dy <- P_0 * exp(-z*L/p) - y*ltlambda
    list(dy)
} 
```

Phase plot at surface
```{r}
#t <- seq(0,40,.5)
nuclide.flow <- flowField(be10, x.lim = c(0,10), y.lim = c(0,10), 
                          system = 'one.dim', parameters = 0, add = F)
nuclide.null <- nullclines(be10, x.lim = c(0,10), y.lim = c(0,10), 
                           system = 'one.dim', points = 200, parameters = 0)
```

Phase plot at depth
```{r}
nuclide.flow <- flowField(be10, x.lim = c(0,10), y.lim = c(0,10), 
                          system = 'one.dim', parameters = 5, add = F)
nuclide.null <- nullclines(be10, x.lim = c(0,10), y.lim = c(0,10), 
                           system = 'one.dim', points = 200, parameters = 5)
```

## Working in GRASS
Use the rgrass7 library to try doing the same thing in GRASS
```{r}
library(rgrass7)
initGRASS(gisBase = "/usr/lib/grass72", home = tempdir(), 
          SG = as(rast, 'SpatialGrid'),
          remove_GISRC = T, override = T)
```

Use r.mapcalc instead of the r function
```{r}
doGRASS('r3.mapcalc', flags = c('overwrite'), expression = expr)
execGRASS('r3.mapcalc', echoCmd = T)
```
```{r}
expr <- paste0(c('test_out = ', P_0, ' * ', 'exp(-1 * z() *', L, ' / ', p,') - N * ', ltlambda), collapse = '')
expr
```
need to work on the z() command
no, actually just use a tibble in r to store the stratigraphy

extra grass commands
need to find the top and bottom values of the dem and set them as the t and b bounds in the 3d region first
g.region t=1305 b=540 res=10 tbres=.1                                           


r.to.rast3elev -l --overwrite input=nuclide_init@PERMANENT elevation=DEM output=nuclide_3d
make a 3d raster from a dem and a data, filling under bounds


Run r.watershed to calculate drainage direction map.
```{r}
doGRASS('r.watershed', flags = c('overwrite'), elevation = dem, drainage = dir) 
```


Select a point

for each time step:
  if erosion rate is positive:
    subtract that depth from nuclide profile
  if deposition rate is positive:
    get drainage direction map
    calculate watershed at that point using r.water.outlet
    get all the cells in that watershed
    add up the nucldie enrichment of all the cells that eroded in the watershed, average based on the depth of erosion
    add that value of nuclide in the depth of deposition  at that area.

                             

r.water.outlet --overwrite input=draindir_45@PERMANENT output=basin_temp coordinates=724673.750147,4284260.05826